Appendix 2: Manuscript and Exploratory Code

### 2025-06-10 manuscript_code.Rmd
changes to factor score graphs

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(worcs)
library(tidyverse)
library(gtsummary)
library(likert)
library(here)
library(corrplot)
library(ggstats)
library(lavaan)
library(lavaanPlot)
library(kableExtra)
library(naniar)
#library(semPlot) # update plots to semPlot later
#library(blavaan) # could not get it to work with this model and synthetic data, try as sensitivity analysis later
#library(tidySEM) # https://github.com/yrosseel/lavaan/issues/359 lavaan does not work if tidySEM is loaded!

@@

#lavaan does not understand empty factor levels
data.fit <- data %>%
  #lavaan cannot deal with unordered factor with more then 2 levels
  filter(de01 == 'man' | de01 == "woman") %>% # EXCLUSION
  filter(bd01 != 'no') %>% # EXCLUSION
  mutate(
    across(isfactor, ~ fct_drop(.x)
    ) %>% 
      mutate(
        bd04 = as.numeric(bd04), # lavaan cannot handle dates
        )
  ) %>% 
  #Original paper used sum scores instead of FA. 
  mutate(
    ss_psycown = 
      across(starts_with("po"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    ss_selfid = 
      across(starts_with("si"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    ss_intention = 
      across(starts_with("it"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    ss_donation = 
      across(all_of(c("bd02", "bd03", "bd04")), ~as.numeric(.)) %>%
      rowSums(na.rm = TRUE)
    )

@@

```{r}
gg_miss_var(
  data.fit2
  )
```

@@

```{r}
gg_miss_upset(data.fit2,nsets = 10,nintersects=40)
```

@@

gg_miss_case_cumsum(
  data.fit2 %>% arrange(dttma)
)

@@

gg_miss_fct(
  x = data.fit %>% select(-dttma,
                          -de01,
                          -de02,
                          -bd02,
                          -bd04,
                          -ss_psycown,
                          -ss_selfid,
                          -ss_donation,
                          ), 
  fct = ss_intention)

@@

data.fit3 <- data.fit2 %>%
  select(-ss_psycown,
         -ss_selfid, 
         -ss_intention,
         -ss_donation)

@@

po01.model <- ' po =~ po01 + po02 + po03
                si =~ si01 + si02 + si03
                it =~ it01 + it02 + it03
                ac =~ ac01 + ac02 + ac03 + ac04
                ai =~ ai01 + ai02 + ai03 + ai04 
                iv =~ iv01 + iv02 + iv03
                bd =~ bd02 + bd03 + bd04 #bd02 is sure to correlate positively

@@

```{r}
#https://stats.stackexchange.com/questions/340857/serial-mediation-in-r-how-to-setup-the-model

@@

@@

```{r}
if (fitmodels) {
  fit.po02ss <- sem(po02ss.model,  
                    estimator = "MLR", 
                    missing = "listwise",
                    data = data.fit2,
                    auto.fix.first = FALSE, 
                    std.lv = TRUE,
                    fixed.x = TRUE

@@

```{r}
lavaan::inspect(fit.po02ss,"r2")
```

```{r}
fit.po02ss.est  <-standardizedSolution(fit.po02ss)
fit.po02ss.est %>% select(-label,se,z) %>%  filter(op == '~')
```

fit.po02ss.est  <-standardizedSolution(fit.po02ss)
fit.po02ss.est %>% 
  filter(abs(est.std) < 0.3) %>% 
  select(-z,-se) %>% 
  arrange(abs(est.std)) %>% 
  # Just the loadings
  filter(op != '~1')

@@

labelsforplottingnull <- c(
  "ss_psycown" = "Psychological ownership",
  "ss_selfid"  = "Self-identity",
  "ss_intention" = "Intention",
  "ss_donation" = "Blood donation history"
)

@@

```{r}
nBoots <- 1000

if (fitmodels) {
  fit.po03ss <- sem(po02ss.model,  
                    estimator = "ML",  
                    missing = "listwise",
                    data = data.fit2, 
                    bootstrap = nBoots,
                    se = "boot",
                    auto.fix.first = FALSE,
                    std.lv = TRUE,
                    fixed.x = TRUE
  )
  models[['fit.po03ss']] <- fit.po03ss
}

lavaan::summary(fit.po03ss,standardized = TRUE)

@@

```{r}
fit.po03ss.best  <-standardizedSolution(fit.po03ss)
fit.po03ss.best %>% select(-label,se,z) %>%  filter(op == '~')
```

```{r}
fit.po03ss.best %>% 
  filter(abs(est.std) < 0.3) %>% 
  select(-z,-se) %>% 
  arrange(abs(est.std)) %>% 
  # Just the loadings
  filter(op != '~1')

@@

```{r}
lavaanPlot(
  model = fit.po03ss, 
   labels = labelsforplottingnull,
  node_options = list(shape = "box", fontname = "Helvetica"), 
  edge_options = list(color = "grey"), 
  coefs = TRUE, 
  stand = TRUE,
  sig=0.001
  ) %>% lavaanPlot::embed_plot_pdf(here("figures/plsssem03.pdf"))

@@

if (fitmodels) {
  fit.po04ss <- sem(
    po02ss.model,  
    estimator = "MLR", 
    data = data.fit2, 
    auto.fix.first = FALSE,
    std.lv = TRUE,
    missing = "fiml",
    fixed.x = FALSE 
  )
  models[['fit.po04ss']] <- fit.po04ss
}

@@

```{r}
fit.po04ss.est  <-standardizedSolution(fit.po04ss)
fit.po04ss.est %>% select(-label,se,z) %>%  filter(op == '~')
```

```{r}
fit.po04ss.est %>% 
  filter(abs(est.std) < 0.3) %>% 
  select(-z,-se) %>% 
  arrange(abs(est.std)) %>% 
  # Just the loadings
  filter(op != '~1')

@@

```{r}
lavaanPlot(
  model = fit.po04ss, 
   labels = labelsforplottingnull,
  node_options = list(shape = "box", fontname = "Helvetica"), 
  edge_options = list(color = "grey"), 
  coefs = TRUE, 
  stand = TRUE,
  sig=0.001
  ) %>% lavaanPlot::embed_plot_pdf(here("figures/plsssem04.pdf"))

@@

if (fitmodels) {
  fit.po05ss <- sem(
    po02ss.model,  
    estimator = "ML", 
    bootstrap = nBoots,
    se = "boot",
    data = data.fit2, 
    auto.fix.first = FALSE,
    std.lv = TRUE,
    missing = "fiml",
    fixed.x = FALSE 
  )
  models[['fit.po05ss']] <- fit.po05ss
}

lavaan::summary(fit.po05ss,standardized = TRUE)

@@

```{r}
fit.po05ss.est  <-standardizedSolution(fit.po05ss)
fit.po05ss.est %>% select(-label,se,z) %>%  filter(op == '~')
```

```{r}
fit.po05ss.est %>% 
  filter(abs(est.std) < 0.3) %>% 
  select(-z,-se) %>% 
  arrange(abs(est.std)) %>% 
  # Just the loadings
  filter(op != '~1')

@@

```{r}
lavaanPlot(
  model = fit.po05ss, 
   labels = labelsforplottingnull,
  node_options = list(shape = "box", fontname = "Helvetica"), 
  edge_options = list(color = "grey"), 
  coefs = TRUE, 
  stand = TRUE,
  sig=0.001
  ) %>% lavaanPlot::embed_plot_pdf(here("figures/plsssem05.pdf"))

@@

po02.model <- ' 
#measurement model
psycown =~ po01 + po02 + po03
selfid =~ si01 + si02 + si03
intention =~ it01 + it02 + it03
blooddonor =~ bd02 + bd03 +bd04 
#bd01 is just used for exclusion of "no"

@@

if (fitmodels) {
  fit.po03 <- sem(po02.model,  
                  estimator = "ML", 
                  data = data.fit2,
                  bootstrap = nBoots,
                  se = "boot",
                  missing = "listwise",
                  auto.fix.first = FALSE,
                  std.lv = TRUE,
                  fixed.x =TRUE
  )
  models[['fit.po03']] <- fit.po03
}

lavaan::summary(fit.po03,standardized = TRUE)

@@

```{r}
fit.po03.best <- parameterEstimates(fit.po03)
fit.po03.best %>% select(-se,-z) %>% filter(op == "~")
```

```{r}
lavaanPlot(
  model = fit.po03, 
   labels = labelsforplotting,
  node_options = list(shape = "box", fontname = "Helvetica"), 
  edge_options = list(color = "grey"), 
  coefs = TRUE, 
  stand = TRUE,
  sig=0.001
  ) %>% lavaanPlot::embed_plot_pdf(here("figures/pllvsem03.pdf"))

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(worcs)
library(tidyverse)
library(gtsummary)
library(likert)
library(here)
library(corrplot)
library(ggstats)
library(lavaan)
library(lavaanPlot)
library(kableExtra)
library(naniar)
library(gridExtra)
#library(semPlot) # update plots to semPlot later
#library(blavaan) # could not get it to work with this model and synthetic data, try as sensitivity analysis later
#library(tidySEM) # https://github.com/yrosseel/lavaan/issues/359 lavaan does not work if tidySEM is loaded!
#library(griExtra) #used for combining multiple plots

@@

#lavaan does not understand empty factor levels
data.fit <- data %>%
  #lavaan cannot deal with unordered factor with more then 2 levels
  filter(de01 == 'man' | de01 == "woman") %>% # EXCLUSION
  filter(bd01 != 'no') %>% # EXCLUSION
  mutate(
    across(isfactor, ~ fct_drop(.x)
    ) %>% 
      mutate(
        bd04 = as.numeric(bd04), # lavaan cannot handle dates
        )
  ) %>% 
  #Original paper used sum scores instead of FA. 
  mutate(
    ss_psycown = 
      across(starts_with("po"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    ss_selfid = 
      across(starts_with("si"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    ss_intention = 
      across(starts_with("it"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    #ss_donation as only measuring bd02
    ss_donation = 
      across(all_of(c("bd02")), ~as.numeric(.)) %>%
      rowSums(na.rm = TRUE)
    )

@@

```{r}
eval=FALSE
gg_miss_var(
  data.fit2
  )
```

@@

```{r}
eval=FALSE
gg_miss_upset(data.fit2,nsets = 10,nintersects=40)
```

@@

```{r}
eval=FALSE
gg_miss_case_cumsum(
  data.fit2 %>% arrange(dttma)
)

@@

```{r}
eval=FALSE
gg_miss_fct(
  x = data.fit %>% select(-dttma,
                          -de01,
                          -de02,
                          -bd02,
                          -bd04,
                          -ss_psycown,
                          -ss_selfid,
                          -ss_donation,
                          ), 
  fct = ss_intention)

@@

```{r}
eval=FALSE
data.fit3 <- data.fit2 %>%
  select(-ss_psycown,
         -ss_selfid, 
         -ss_intention,
         -ss_donation)

@@

```{r}
#making bd02 the only variable in donation history

po01.model <- ' po =~ po01 + po02 + po03
                si =~ si01 + si02 + si03
                it =~ it01 + it02 + it03
                ac =~ ac01 + ac02 + ac03 + ac04
                ai =~ ai01 + ai02 + ai03 + ai04 
                iv =~ iv01 + iv02 + iv03
                bd =~ bd02

@@

```{r}
eval=FALSE
#https://stats.stackexchange.com/questions/340857/serial-mediation-in-r-how-to-setup-the-model

@@

```
```{r}
#Bivariate correlation table for paper

@@

```{r}
eval=FALSE
if (fitmodels) {
  fit.po02ss <- sem(po02ss.model,  
                    estimator = "MLR", 
                    missing = "listwise",
                    data = data.fit2,
                    auto.fix.first = FALSE, 
                    std.lv = TRUE,
                    fixed.x = TRUE

@@

```{r}
eval=FALSE
lavaan::inspect(fit.po02ss,"r2")
```

```{r}
eval=FALSE
fit.po02ss.est  <-standardizedSolution(fit.po02ss)
fit.po02ss.est %>% select(-label,se,z) %>%  filter(op == '~')
```

```{r}
eval=FALSE
fit.po02ss.est  <-standardizedSolution(fit.po02ss)
fit.po02ss.est %>% 
  filter(abs(est.std) < 0.3) %>% 
  select(-z,-se) %>% 
  arrange(abs(est.std)) %>% 
  # Just the loadings
  filter(op != '~1')

@@

```{r}
eval=FALSE
labelsforplottingnull <- c(
  "ss_psycown" = "Psychological ownership",
  "ss_selfid"  = "Self-identity",
  "ss_intention" = "Intention",
  "ss_donation" = "Blood donation history"
)

@@

```{r}
eval=FALSE
#nBoots <- 1000

#if (fitmodels) {
  #fit.po03ss <- sem(po02ss.model,  
                    #estimator = "ML",  
                    #missing = "listwise",
                    #data = data.fit2, 
                    #bootstrap = nBoots,
                    #se = "boot",
                    #auto.fix.first = FALSE,
                    #std.lv = TRUE,
                    #fixed.x = TRUE
  #)
  #models[['fit.po03ss']] <- fit.po03ss
#}

#lavaan::summary(fit.po03ss,standardized = TRUE)

@@

```{r}
#eval=FALSE
#fit.po03ss.best  <-standardizedSolution(fit.po03ss)
#fit.po03ss.best %>% select(-label,se,z) %>%  filter(op == '~')
```

```{r}
eval=FALSE
#fit.po03ss.best %>% 
  #filter(abs(est.std) < 0.3) %>% 
  #select(-z,-se) %>% 
  #arrange(abs(est.std)) %>% 
  # Just the loadings
  #filter(op != '~1')

@@

```{r}
eval=FALSE
#lavaanPlot(
  #model = fit.po03ss, 
   #labels = labelsforplottingnull,
  #node_options = list(shape = "box", fontname = "Helvetica"), 
  #edge_options = list(color = "grey"), 
  #coefs = TRUE, 
  #stand = TRUE,
  #sig=0.001
  #) %>% lavaanPlot::embed_plot_pdf(here("figures/plsssem03.pdf"))

@@

```{r}
eval=FALSE
if (fitmodels) {
  fit.po04ss <- sem(
    po02ss.model,  
    estimator = "MLR", 
    data = data.fit2, 
    auto.fix.first = FALSE,
    std.lv = TRUE,
    missing = "fiml",
    fixed.x = FALSE 
  )
  models[['fit.po04ss']] <- fit.po04ss
}

@@

```{r}
eval=FALSE
fit.po04ss.est  <-standardizedSolution(fit.po04ss)
fit.po04ss.est %>% select(-label,se,z) %>%  filter(op == '~')
```

```{r}
eval=FALSE
fit.po04ss.est %>% 
  filter(abs(est.std) < 0.3) %>% 
  select(-z,-se) %>% 
  arrange(abs(est.std)) %>% 
  # Just the loadings
  filter(op != '~1')

@@

```{r}
eval=FALSE
lavaanPlot(
  model = fit.po04ss, 
   labels = labelsforplottingnull,
  node_options = list(shape = "box", fontname = "Helvetica"), 
  edge_options = list(color = "grey"), 
  coefs = TRUE, 
  stand = TRUE,
  sig=0.001
  ) %>% lavaanPlot::embed_plot_pdf(here("figures/plsssem04.pdf"))

@@

```{r}
eval=FALSE
#if (fitmodels) {
  #fit.po05ss <- sem(
    #po02ss.model,  
    #estimator = "ML", 
    #bootstrap = nBoots,
    #se = "boot",
    #data = data.fit2, 
    #auto.fix.first = FALSE,
    #std.lv = TRUE,
    #missing = "fiml",
    #fixed.x = FALSE 
  #)
  #models[['fit.po05ss']] <- fit.po05ss
#}

#lavaan::summary(fit.po05ss,standardized = TRUE)

@@

```{r}
eval=FALSE
#fit.po05ss.est  <-standardizedSolution(fit.po05ss)
#fit.po05ss.est %>% select(-label,se,z) %>%  filter(op == '~')
```

```{r}
eval=FALSE
#fit.po05ss.est %>% 
  #filter(abs(est.std) < 0.3) %>% 
  #select(-z,-se) %>% 
  #arrange(abs(est.std)) %>% 
  # Just the loadings
  #filter(op != '~1')

@@

```{r}
eval=FALSE
#lavaanPlot(
  #model = fit.po05ss, 
   #labels = labelsforplottingnull,
  #node_options = list(shape = "box", fontname = "Helvetica"), 
  #edge_options = list(color = "grey"), 
  #coefs = TRUE, 
  #stand = TRUE,
  #sig=0.001
  #) %>% lavaanPlot::embed_plot_pdf(here("figures/plsssem05.pdf"))

@@

#making bd02 the only variable in donation history

po02.model <- ' 
#measurement model
psycown =~ po01 + po02 + po03
selfid =~ si01 + si02 + si03
intention =~ it01 + it02 + it03
blooddonor =~ bd02 
#bd01 is just used for exclusion of "no"

@@

#if (fitmodels) {
  #fit.po03 <- sem(po02.model,  
                  #estimator = "ML", 
                  #data = data.fit2,
                  #bootstrap = nBoots,
                  #se = "boot",
                  #missing = "listwise",
                  #auto.fix.first = FALSE,
                  #std.lv = TRUE,
                  #fixed.x =TRUE
  #)
 # models[['fit.po03']] <- fit.po03
#}

#lavaan::summary(fit.po03,standardized = TRUE)

@@

```{r}
#fit.po03.best <- parameterEstimates(fit.po03)
#fit.po03.best %>% select(-se,-z) %>% filter(op == "~")
```

```{r}
#lavaanPlot(
  #model = fit.po03, 
   #labels = labelsforplotting,
  #node_options = list(shape = "box", fontname = "Helvetica"), 
  #edge_options = list(color = "grey"), 
  #coefs = TRUE, 
  #stand = TRUE,
  #sig=0.001
  #) %>% lavaanPlot::embed_plot_pdf(here("figures/pllvsem03.pdf"))

@@

#Exploratory Analysis (a1*b1 comparison)

factor_scores <- lavPredict(fit.po04, type = "lv")
head(factor_scores)
summary(factor_scores)

df_scores <- cbind(
  as.data.frame(factor_scores),
  de02 = data.fit2$de02, #de02 = age
  de01 = data.fit2$de01 #de01 = gender
)

df_scores$de01 <- factor(df_scores$de01, levels = c(1, 2), labels = c("man", "woman"), ordered = TRUE)

#psycown ~ blooddonor
factor1 <- ggplot(df_scores, aes(x = psycown, y = blooddonor)) +
  geom_point(aes(color = de02), alpha = 0.7, size = 2.5) +
  geom_smooth(aes(linetype = de01), colour = "black", se = FALSE) +
  labs(title = "Factor scores: ownership x identity",
       x = "Identity",
       y = "Donation history",
       color = "Age",
       linetype = "Gender") +
  theme_minimal()

#selfid ~ blooddonor
factor2 <- ggplot(df_scores, aes(x = selfid, y = blooddonor)) +
  geom_point(aes(color = de02), alpha = 0.7, size = 2.5) +
  geom_smooth(aes(linetype = de01), colour = "black", se = FALSE) +
  labs(title = "Factor scores: ownership x identity",
       x = "Identity",
       y = "Donation history",
       color = "Age",
       linetype = "Gender") +
  theme_minimal()

#selfid ~ psycown
factor3 <- ggplot(df_scores, aes(x = psycown, y = selfid)) +
  geom_point(aes(color = de02), alpha = 0.7, size = 2.5) +
  geom_smooth(aes(linetype = de01), colour = "black", se = FALSE) +
  labs(title = "Factor scores: ownership x identity",
       x = "Psychological Ownership",
       y = "Identity",
       color = "Age",
       linetype = "Gender") +
  theme_minimal()

#intention ~ blooddonor
factor4 <- ggplot(df_scores, aes(x = intention, y = blooddonor,
                      color = as.factor(de01))) +
  geom_point(aes(color = de02), alpha = 0.7, size = 2.5) +
  geom_smooth(aes(linetype = de01), colour = "black", se = FALSE) +
  labs(title = "Factor scores: ownership x identity",
       x = "Intention",
       y = "Donation History",
       color = "Age",
       linetype = "Gender") +
  theme_minimal()

#intention ~ psycown
factor5 <- ggplot(df_scores, aes(x = psycown, y = intention,
                      color = as.factor(de01))) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_smooth() +
  labs(title = "Factor Scores: ownership x intention",
       x = "Psychological Ownership",
       y = "Intention",
       color = "Age",
       linetype = "Gender") +
  theme_minimal()

#intention ~ selfid
factor6 <- ggplot(df_scores, aes(x = psycown, y = intention,
                      color = as.factor(de01))) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_smooth() +
  labs(title = "Factor Scores: ownership x intention",
       x = "Psychological Ownership",
       y = "Intention",
       color = "Age",
       linetype = "Gender") +
  theme_minimal()

combined <- grid.arrange(factor1, factor2, factor3, factor4, factor5, factor6, ncol = 2)

ggsave(
  "Ownership and Intention Factor Loadings",
  combined, width = 20, 
  height = 10, 
  device = "pdf")

#exploratory analysis (multigroup analysis to explore impact of gender)

#making gender a factor and ordering

data.fit2$de01 <- factor(data.fit2$de01, levels = c(1, 2), labels = c("man", "woman"), ordered = TRUE)

data.fit2 %>%
  group_by(de01) %>%
  summarise(across(everything(), ~ var(.x, na.rm = TRUE)))

#defining new model that doesn't contained de01

ea2.model <- ' 
#measurement model
psycown =~ po01 + po02 + po03
selfid =~ si01 + si02 + si03
intention =~ it01 + it02 + it03
blooddonor =~ bd02

# regressions (separate labels per group)
psycown ~ c(a1g1, a1g2) * blooddonor
selfid ~ c(a2g1, a2g2) * blooddonor + c(d21g1, d21g2) * psycown
intention ~ c(cpg1, cpg2) * blooddonor + c(b1g1, b2g2) * psycown + c(b2g1, b2g2) * selfid

#g1 indirect effects
ind_eff_g1 := a1g1 * d21g1 * b2g1
ind_eff1_g1:= a1g1 * b1g1
ind_eff2_g1:= a2g1 * b2g1

#g2 indirect effects
ind_eff_g2:= a1g2 * d21g2 * b2g2
ind_eff1_g2:= a1g2 * b1g2
ind_eff2_g2:= a2g2 * b2g2
'

ea2.free <- sem(
  ea2.model,
  data = data.fit2,
  group = "de01",
  estimator = "MLR",
  std.lv = TRUE,
  missing = "fiml",
  fixed.x = FALSE,
  auto.fix.first = FALSE,
  group.equal = c("loadings","intercepts”)
)

ea2.fixed <- sem(
ea2.model,
data = data.fit2,
group = "de01",
estimator = "MLR",
std.lv = TRUE,
missing = "fiml",
fixed.x = FALSE,
auto.fix.first = FALSE,
group.equal = c("loadings", ”intercepts”, ”regressions”),
)

# this is for model comparison
lavTestLRT(ea2.free, ea2.fixed)

#ea2 <- sem(
  #ea2.model,
  #data = data.fit2,
  #group = "de01",
  #estimator = "MLR",
  #std.lv = TRUE,
  #missing = "fiml",
  #fixed.x = FALSE,
  #auto.fix.first = FALSE
#)

#summary(ea2, standardized = TRUE)

#end of exploratory analysis
knitr::knit_exit()
